
## Introduction to R

<img align="right" src=images/R_logo.png width="150">

- Date: 15th March 2018, 6 - 7pm
- Series: Wolfson College [Skills for Academic Success](https://www.wolfson.cam.ac.uk/study-skills)
- Location: [Roger Needham Room](http://www.wolfson.cam.ac.uk/tour/chancellorscentre), [Wolfson college](https://goo.gl/maps/aR6a5FWrLoR2), University of Cambridge, UK
- Register [here](https://www.eventbrite.co.uk/e/skills-for-academic-success-introduction-to-r-tickets-41949950431)
- Trainer: Sergio Martínez Cuesta


### Overview

This course provides a short beginners introduction to the R programming language and software environment for statistical computing and graphics. Sergio will demonstrate basic examples on how to input, explore, plot and output data in R. Everybody is welcome, if you would like to follow along with your laptop, please bring R and RStudio [downloaded and installed](https://www.rstudio.com/products/rstudio/download/) before the session.



### Outline

- Motivation
- Installing R and RStudio
- How can I find help?
- Getting started
- Variables and functions
    - Exercise 1

This short course is based on the [R crash course](https://github.com/bioinformatics-core-shared-training/r-crash-course) developed by Mark Dunning and Laurent Gatto.



### Motivation

- R is one of the most widely-used programming languages for data analysis and visualisation in academia and industry
- It is open-source and available in all platforms (Mac, Linux and Windows)
- Supported by a broad community of software developers and researchers who contribute R packages and libraries to many fields of research 
- It facilitates reproducibility in research and integration of all your analyses in individual scripts
- Easy to write documentation and code together using a free environment like [RStudio](https://www.rstudio.com/)

E.g. The New Zealand [Tourism Dashboard](https://mbienz.shinyapps.io/tourism_dashboard_prod/) uses R extensively to report statistics.



### Installing R and RStudio

- [Latest version of R](https://cran.r-project.org/)
    - To check if you have R installed in Mac, go to `Finder` -> `Applications` -> `Utilities` -> `Terminal` and type `R`
- [RStudio Desktop Open Source License](https://www.rstudio.com/products/rstudio/download/)
    - To check if you have RStudio installed in Mac, go to `Finder` -> `Applications` and click on `Rstudio` 



### How can I find help?

- [Stack Overflow](https://stackoverflow.com/)
- [CRAN Task Views](https://cran.r-project.org/web/views/)
- [R-bloggers](https://www.r-bloggers.com/)
- [Quick-R](https://www.statmethods.net/)
- Local R groups, e.g. [R-ladies Cambridge](https://www.meetup.com/rladies-cambridge/)
- Type `?` followed my the name of the function that you'd like to use, e.g. `?mean`

If you are interested in bioinformatics and computational biology, the following links might also be of interest:

- [Bioconductor](https://www.bioconductor.org/)
- [Biostars](https://www.biostars.org/)



### Getting started

- We will be using RStudio console (bottom-left panel) to interact with R during the workshop
- The blocks of code shown below are written using the format **R markdown**, which allows mixed plain text and R code together in the same document
- Each line of R code within a block can be executed by clicking on the line and pressing CMD + ENTER (Mac) or CTRL + ENTER (Windows and Linux). Alternatively, to execute the entire block, click on the green arrow tip on the right-hand side of the block. Explore with the following:

```{r eval=FALSE}
print("R is fun!")
```

```{r eval=FALSE}
3 + 1
```

- You can add a new block of code by selecting `R` in the `Insert` menu or by typing the following syntax directly:

```{r}
# R code goes here
```



### Variables and functions

You can use R as a calculator using the symbols `+`, `-`, `*` and `/`, or more advanced such as statistical operations, logarithms, trigonometry ...

```{r eval=FALSE}
2 + 1
7 - 1
3 * 2
10 / 5

mean(1:5)
log(1)
pi
sin(pi/2)
```

To store your results for later, use **variables**. To create them, use the assignment operator `<-`:

```{r eval=FALSE}
x <- 25
x
y <- 16
y
```

You can perform multiple operations using variables:

```{r eval=FALSE}
sqrt(x)
x + y
x <- 36
x <- y
x <- x + 8
```

**Functions** in R take one or more *arguments* as input, which are captured using parentheses. Arguments can be named explicitly, otherwise they are meant to be used in the same order as described in the function definition. E.g. `seq` is a function for generating a numeric sequence *from* and *to* particular numbers. Type `?seq` to get the help page for this function.

```{r eval=FALSE}
?seq
seq(from = 1, to = 10, by = 2)
seq(1, 10, 2)
```

Some functions have *default* values:

```{r eval=FALSE}
seq(1, 10, 1)
seq(1, 10)
```



#### Exercise 1




### Additional materials

Blogs:

- [End-to-end visualization using ggplot2](https://rviews.rstudio.com/2017/08/14/end-to-end-visualization-using-ggplot2/)
- [ggplot2 - Easy way to mix multiple graphs on the same page](http://www.sthda.com/english/wiki/ggplot2-easy-way-to-mix-multiple-graphs-on-the-same-page)
- [Getting started with data visualization in R using ggplot2](http://www.storybench.org/getting-started-data-visualization-r-using-ggplot2/)

Books:

- [R for Data Science](http://r4ds.had.co.nz/)
- [Data Visualization for Social Science. A practical introduction with R and ggplot2](http://socviz.co/)

Courses:

- [CRUK-CI R crash course](https://bioinformatics-core-shared-training.github.io/r-crash-course/)
- [R for Reproducible Scientific Analysis](http://swcarpentry.github.io/r-novice-gapminder/)
- [Karl Broman's mini tutorials](http://kbroman.org/pages/tutorials.html)
- [Basic statistics and data handling with R](https://github.com/cambiotraining/stats-intro)
- [Scripting for data analysis (with R)](https://github.com/mrtnj/scripting_for_data_analysis)
- [An Introduction to Solving Biological Problems with R](http://cambiotraining.github.io/r-intro/)
- [Data Analysis and Visualisation using R](http://bioinformatics-core-shared-training.github.io/r-intermediate/): including dplyr and ggplot2
- Babraham institute basic/advanced R and ggplot2 [courses](http://www.bioinformatics.babraham.ac.uk/training/)
- R object-oriented programming and package development, [link1](http://lgatto.github.io/TeachingMaterial/) and [link2](http://logic.sysbiol.cam.ac.uk/teaching/advancedR/)
- [R course content for the CODATA-RDA Research Data Science Summer School](https://github.com/marioa/trieste)
- [Data carpentry course for biologists](https://jabberwocky.weecology.org/2016/11/14/fork-our-course-a-semester-long-data-carpentry-course-for-biologists/) by Ethan White
- [Cambridge's Data carpentry using R](https://tavareshugo.github.io/2017-09-11-cambridge/)

Perspectives:

- [Bioconductor: A test drive of a DNA-analysis toolkit in the cloud](https://www.nature.com/articles/d41586-017-07833-1)
- [Scientific computing: code alert](https://www.nature.com/nature/journal/v541/n7638/full/nj7638-563a.html)


### TODO
